---
title: "eechidna: An R package for extracting, tidying and combining data from Australian Elections and Censuses"
author:
  - name: Ben Marwick
    affiliation: University of Washington
    address: USA
    email: bmarwick@uow.edu.au
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    - Clayton, VIC 3800, Australia
    email:  dicook@monash.edu
  - name: Anthony Ebert
    affiliation: Queensland University of Technology
    address: Australia
    email: anthonyebert@gmail.com
  - name: Heike Hofmann
    affiliation: Iowa State University
    address: USA
    email: hofmann@iastate.edu
  - name: Rob J Hyndman
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    - Clayton, VIC 3800, Australia
    email:  Rob.Hyndman@monash.edu
  - name: Thomas Lumley
    affiliation: University of Auckland
    address: New Zealand
    email: t.lumley@auckland.ac.nz
  - name: Carson Sievert
    affiliation: Plotly
    address: USA
    email: cpsievert1@gmail.com
  - name: Mingzhu Sun
    affiliation: University of Queensland
    address: Australia
    email: m.sun1@uq.edu.au
  - name: Dilini Talagala
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    - Clayton, VIC 3800, Australia
    email: Dilini.Talagala@monash.edu
  - name: Nicholas Tierney
    affiliation: Queensland University of Technology
    address: Australia
    email: nicholas.tierney@gmail.com
  - name: Nathaniel Tomasetti
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    - Clayton, VIC 3800, Australia
    email: Nathaniel.Tomasetti@monash.edu
  - name: Earo Wang
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    - Clayton, VIC 3800, Australia
    email:  Earo.Wang@monash.edu
  - name: Fang Zhou
    affiliation: Microsoft
    address: Singapore
    email: Fang.Zhou@microsoft.com

abstract: >
  An abstract of less than 150 words. TBC.
bibliography: RJreferences.bib
output: rticles::rjournal_article
preamble:
  \input preamble.tex
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, messages=FALSE, warning=FALSE,fig.width=5.2,fig.height=5.2/1.6)
```

```{r packages, echo=FALSE, cache=FALSE, message=FALSE}
# Required packages to reproduce this analysis
library(eechidna)
library(ggplot2)
library(dplyr)
library(purrr)
library(knitr)
library(broom)
library(tidyr)
library(ggthemes)
library(scales)
library(gridExtra)
library(GGally)
library(corrplot)
```


#Introduction


The `eechidna` package for R \citep{R} provides data from the 2013 and 2016 Australian Federal Elections and the 2011 and 2016 Australian Censuses for each House of Representatives electorate, along with some tools for visualizing and analysing the data. The name of the package is an acronym for ``Exploring Election and Census Highly Informative Data Nationally for Australia'', and is an allusion to the echidna, an Australian egg-laying marsupial.

This package was initially developed during the rOpenSci auunconf event in Brisbane, Queensland, during 21--22 April 2016. It was later updated during the BURGr R Unconference in Brisbane on 31 March--1 April 2017.

The `nzelect` \citep{nzelect} package was an important inspiration for this pack


#Data

## Election Data

The data for the 2013 and 2016 Australian Federal Elections for the House of Representative were obtained from \url{http://results.aec.gov.au/}. These data are copyright Commonwealth of Australia (Australian Electoral Commission) 2013, 2016.

The election results are made available in csv format. There are many different summaries which matches the complexity of the Australian system that has a majority-preferential instant-runoff voting in the lower house, a single transferable vote proportional representation system in the upper house, and compulsory voting. In the `eechidna` package we have eight election datasets:

\begin{table}[htb]\centering\small
\begin{tabular}{ll}
\toprule
\bfseries Data object name & \bf Description \\
\midrule
\bfseries 2013 election\\ 
\verb|aec2013_fp|              & first preference votes for candidates at each polling place\\
\verb|aec2013_fp_electorate|   & first preference votes for candidates at each electorate\\
\verb|aec2013_2pp|             & two party preferred votes for candidates at each polling place\\
\verb|aec2013_2pp_electorate|  & two party preferred votes for candidates at each electorate\\
\bfseries 2016 election\\ 
\verb|aec2016_fp|              & first preference votes for candidates at each polling place\\
\verb|aec2016_fp_electorate|   & first preference votes for candidates at each electorate\\
\verb|aec2016_2pp|             & two party preferred votes for candidates at each polling place\\
\verb|aec2016_2pp_electorate|  & two party preferred votes for candidates at each electorate\\
\bottomrule
\end{tabular}
\caption{Australian Election Data sets.}
\end{table}

The 2013 data sets contains `r formatC(nrow(aec2013_fp), big.mark=",", drop0trailing=TRUE, format="f")` rows and `r ncol(aec2013_fp)` columns. The 2016 data sets contains `r formatC(nrow(aec2016_fp), big.mark=",", drop0trailing=TRUE, format="f")` rows and `r ncol(aec2016_fp)` columns. Note that some parties have slightly different names in different states. For example there is "The Nationals" and the "National Party". We've simplified the data by renaming some of the parties to the variant of the name that won the most votes. 

\todo[inline]{Say something about processing done to get data in this form.}

## Election Maps

Spatial format files are used to distribute the the geographic extents of the electorates. The formats provided are ESRI and MapInfo which can be imported into R using a variety of tools. Maps are provided on the national, and also state level, and it is clear that the states are responsible for managing their electorate boundaries because there is some variation in formats and timing of release from state to state. These files allow the election map of the country to be drawn. The electorates do change from one election cycle to another. For example, between 2013 and 2016, Western Australia gained an electorate, NSW lost one, resulting in a renaming also, and the ACT changed and renamed one electorate. In addition, two electorates had their names spelled differently between the two years, Mcmillan to McMillan and Mcpherson to McPherson.

```{r echo=FALSE}
data(nat_map)
# choropleth map with Census data
nat_map$region <- nat_map$ELECT_DIV
data(abs2011)
abs2011$region <- abs2011$Electorate
library(ggplot2)
library(ggthemes)
both <- intersect(unique(abs2011$region), unique(nat_map$region))
ggplot(aes(map_id=region), data=subset(abs2011, region %in% both)) +
  geom_map(aes(fill=MedianIncome), map=subset(nat_map, region %in% both)) +
  expand_limits(x=nat_map$long, y=nat_map$lat) + 
  coord_map() +
  theme_map() + theme(legend.position="right")
```

## Census Data

The 2011 and 2016 Australian Census data were obtained from the Australian Bureau of Statistics and are available in the package as `abs2011` and `abs2016`. These data frames contain 150 rows each, corresponding to the 150 electorates in the 2013 and 2016 Australian House of Representatives elections (\url{http://www.aec.gov.au/profiles/}). The `r ncol(abs2011)` columns in each data frame largely comprise demographic summaries of each electorate. These are summarized in Table \ref{abs2011}.


```{r make-data-dictionary, include = FALSE}
data_dictionary <- data.frame(
  Variable = paste("\\",colnames(abs2011),"|",sep=""),
  Details = c(
    "Commonwealth Electoral District identifier",
    "Name of electorate",
    "State containing electorate",
    "Total population of electorate",
    "Area of electorate in square kilometres",
    "Median income of people within electorate",
    "Percentage of people unemployed",
    "Percentage of people whose highest qualification is a Bachelor degree",
    "Percentage of people whose highest qualification is a postgraduate degree",
    "Percentage of people affiliated with the Christian religion (of all denominations)",
    "Percentage of people affiliated with the Catholic denomimation.",
    "Percentage of people affiliated with the Buddhist religion.",
    "Percentage of people affiliated with the Islam religion.",
    "Percentage of people affiliated with the Jewish religion. ",
    "Percentage of people with no religion.",
    "Percentage of people aged 0-4.",
    "Percentage of people aged 5-9.",
    "Percentage of people aged 15-19.",
    "Percentage of people aged 20-24.",
    "Percentage of people aged 25-34.",
    "Percentage of people aged 35-44.",
    "Percentage of people aged 45-54.",
    "Percentage of people aged 55-64.",
    "Percentage of people aged 65-74.",
    "Percentage of people aged 75-84.",
    "Percentage of people aged 85 or higher.",
    "Percentage of people born outside Australia.",
    "Percentage of people who are Indigenous",
    "Percentage of people who speak only English",
    "Percentage of people who speak a language other than English at home",
    "Percentage of people who are married",
    "Percentage of people who are in a de facto marriage",
    "Total number of families to total number of people (times 100)",
    "Percentage of people with home internet",
    "Percentage of dwellings not owned (either outright or with a mortgage)")
) # close data_frame
```

```{r display-data-dictionary, cache=FALSE, results="asis", message=FALSE}
xtable::xtable(data_dictionary,
               label="abs2011",
               caption="Variables from the 2011 Census Data.") %>%
print(type="latex",
  size="footnotesize",
  comment=FALSE,
  booktabs=TRUE,
  include.rownames=FALSE,
  sanitize.colnames.function =  function(x){paste0('\\textbf{', x, '}')
},
  sanitize.text.function =
function(str) gsub("\\", "\\verb|", str, fixed = TRUE))
```



\todo[inline]{Say something about processing done to get data in this form.}



#Exploring the 2011 Census data

There are many age groups. To look at all of them at once, we can gather them into a dataframe ready for plotting using `tidyr`.

```{r abs2011, fig.cap="Test", echo=TRUE}
abs2011 %>%
  select(starts_with("Age"), 
         Electorate) %>%
  gather(key = "Age",
         value = "Percent_in_electorate",
         -Electorate) %>% 
  ggplot(data = .,
         aes(x = reorder(Age, - Percent_in_electorate),
             y = Percent_in_electorate,
             colour = Age)) +
  geom_boxplot() + 
  coord_flip() + 
  theme_minimal() + 
  theme(legend.position = "none") +
  labs(x = "Age Groups",
       y = "% in Electorate")
```

We can look at the relationship between income and state (Figure \ref{fig:abs2011b})  and the relationship between income and education (Figure \ref{fig:abs2011c}).


```{r abs2011b, fig.cap="TBC"}
ggplot(data = abs2011,
       aes(x = reorder(State, -MedianIncome),
           y = MedianIncome,
           colour = State)) + 
  geom_boxplot() + 
  geom_jitter(alpha = 0.35, 
              size = 2,
              width = 0.1) +
  theme_minimal() + 
  theme(legend.position = "none") + 
  labs(x = "State") +
  coord_flip()
```


```{r abs2011c, fig.cap="TBC", message=FALSE}
abs2011 %>%
  select(Postgraduate,
         Bachelor,
         MedianIncome) %>% 
  gather(key = "Education",
         value = "Prop_Educated",
         -MedianIncome) %>%
ggplot(data = ,
       aes(x = Prop_Educated,
           y = MedianIncome,
           colour = Education)) + 
  geom_point() + 
  geom_smooth() +
  theme_minimal() +
  scale_color_brewer(type = "qual", palette = "Set1")
```

Let's look at all of the religions

```{r abs2011d, fig.cap="TBC"}
abs2011 %>%
  select(Christianity,
         Catholic,
         Buddhism,
         Islam,
         Judaism,
         NoReligion) %>%
  gather(key = "ReligionType",
         value = "Percent") %>%
  ggplot(data = .,
         aes(x = reorder(ReligionType, -Percent),
             y = Percent,
             colour = ReligionType)) + 
  geom_boxplot() + 
  theme_minimal() + 
  theme(legend.position = "none") +
  coord_flip() + 
  labs(x = "Religion")
```



```{r}
ggplot(data = abs2011,
       aes(x = reorder(State, -Christianity),
           y = Christianity,
           colour = State)) + 
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_flip() + 
  labs(x = "State")
  
           

```

## Internet by state

```{r}

ggplot(data = abs2011,
       aes(x = reorder(State, -Internet),
           y = Internet,
           colour = State)) + 
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_flip() + 
  labs(x = "State")

```

#Exploring the 2016 Election data

We can summarise the data to discover some basic details about the 2016 election. Let's start by reproducing the overall result of the election by finding out which party won the most electorates according to the two party preferred votes:

```{r who_won}
who_won <- aec2016_2pp_electorate %>% 
  group_by(PartyNm) %>% 
  tally() %>% 
  arrange(desc(n)) 
ggplot(who_won, 
       aes(reorder(PartyNm, n), 
           n)) +
  geom_point(size = 2) + 
  coord_flip() + 
  scale_y_continuous(labels = comma) +
  theme_bw() +
  ylab("Total number of electorates") +
  xlab("Party") +
  theme(text = element_text(size=10))
```

This result above matches the official AEC result, showing the Liberal party winning the most electorates. 

## Which party received the most ordinary votes?

An alternative way to evaluate the outcome of the election is by counting the number of ordinary first preference votes for each party (not including postal votes, preference flows, etc.). Here we can find the total number of ordinary votes for each party:

```{r total_votes_for_parties}
total_votes_for_parties <- aec2016_fp %>% 
  dplyr::select(PartyNm, OrdinaryVotes) %>% 
  dplyr::group_by(PartyNm) %>% 
  dplyr::summarise(total_votes = sum(OrdinaryVotes, na.rm = TRUE)) %>% 
  ungroup() %>%
  arrange(desc(total_votes))
```

We can plot the ordinary vote count, showing the Liberal Party receiving the most votes.

```{r plot_total_votes_for_parties,  fig.height = 6}
ggplot(total_votes_for_parties, 
       aes(x=reorder(PartyNm, total_votes), 
           y=total_votes)) +
  geom_point(size = 2) + 
  coord_flip() + 
  scale_y_continuous(labels = comma) +
  theme_bw() +
  ylab("Total ordinary votes") +
  xlab("Party") +
  theme(text = element_text(size=10))
```


## Which candidate did the best?

There are a few ways we can answer this question. We can see who received the highest number of ordinary votes, like this:

```{r who_most-votes, results="asis"}
who_most_votes <- aec2013_fp %>% 
  dplyr::filter(CandidateID != 999) %>% #exclude informal votes
  dplyr::mutate(candidate_full_name = paste0(GivenNm, " ", Surname, " (", CandidateID, ")")) %>% 
  dplyr::group_by(candidate_full_name) %>% 
  dplyr::summarise(total_votes_for_candidate = sum(OrdinaryVotes, rm.na = TRUE)) %>% 
  arrange(desc(total_votes_for_candidate))
# inspect
who_most_votes %>% 
  mutate(total_votes_for_candidate = 
           prettyNum(total_votes_for_candidate, 
                     big.mark = ","))  %>% 
  head -> tab
  colnames(tab) <- c("Candidate full name", "Total votes")
  print(xtable::xtable(tab), comment=FALSE)
```

But this is a crude measure of how successful a candidate was in the election, because electorates with more people will result in more votes for the winner, even if the victory is only by a tiny margin. So let's see remove the effect of the population size in the electorates by finding out who won the highest proportion of votes in their electorate. 

```{r who_highest_proportion-votes, results="asis"}
who_most_votes_prop <- aec2013_fp %>% 
  dplyr::filter(CandidateID != 999) %>% #exclude informal votes
  dplyr::mutate(candidate_full_name = 
           paste0(GivenNm, " ", Surname, " (", CandidateID, ")")) %>% 
  dplyr::group_by(Electorate, candidate_full_name) %>% 
  dplyr::summarise(sum_votes = sum(OrdinaryVotes))  %>% 
  dplyr::mutate(prop_votes = round(sum_votes / sum(sum_votes), 3),
         sum_votes = prettyNum(sum_votes, ",")) %>% 
  ungroup %>% 
  arrange(desc(prop_votes))

# inspect
who_most_votes_prop %>% 
  data.frame %>% 
  head -> z
colnames(z) <- c("Electorate","Candidate","Total votes","Proportion votes")
print(xtable::xtable(z), comment=FALSE)
```

In the above table (which shows just the first ten rows for convienience) we see that Alex Hawke received the highest proportion of the vote in any electorate, as well as a few more familiar names doing very well in their electorates. We can see that taking about 65% of the vote in an electorate counts as a huge win. But what is the smallest winning margin that occured in the 2013 election? Let's find the candidates who won their electorate by the slimmest advantage over their competitors. 

## Which candidate won with the smallest number of votes?


```{r who_won_least_votes_prop, results="asis", echo=FALSE}
who_won_least_votes_prop <- aec2016_fp %>% 
   dplyr::filter(CandidateID != 999) %>% # keep only the winners
   dplyr::mutate(candidate_full_name = 
            paste0(GivenNm, " ", Surname, " (", CandidateID, ")")) %>% 
   dplyr::group_by(Electorate, candidate_full_name) %>% 
   dplyr::summarise(sum_votes = sum(OrdinaryVotes)) %>% 
   dplyr::mutate(prop_votes = round(sum_votes / sum(sum_votes), 2)) %>% 
   ungroup %>% 
   dplyr::left_join(aec2013_2pp_electorate %>% 
               dplyr::mutate(candidate_full_name = 
            paste0(GivenNm, " ", Surname, " (", CandidateID, ")")), "candidate_full_name") %>% 
   dplyr::filter(Elected == "Y") %>% 
   dplyr::select(Electorate.x, 
          candidate_full_name, 
          prop_votes, 
          PartyNm) %>% 
   arrange(prop_votes)

# have a look
who_won_least_votes_prop %>%
 head -> z
colnames(z) <- c("Electorate","Candidate","Proportion votes")
print(xtable::xtable(z), comment=FALSE)
```

The table above shows that Clive Palmer won his electorate by the smallest margin, just 27% of the ordinary votes. Bob Katter also won with a very small margin, 29% of the first preferences. Let's see the distribution of votes in Fairfax to see a bit more detail:


```{r fairfax, results="asis"}
fairfax <- aec2016_fp %>% 
  dplyr::filter(CandidateID != 999) %>% # exclude informal votes
  dplyr::filter(Electorate == "Fairfax") %>% 
  dplyr::mutate(candidate_full_name = 
           paste0(GivenNm, " ", 
                  Surname, 
                  " (", 
                  CandidateID, 
                  ")")) %>% 
  dplyr::group_by(Electorate, candidate_full_name) %>% 
  dplyr::summarise(sum_votes = sum(OrdinaryVotes)) %>% 
  dplyr::mutate(prop_votes = round(sum_votes / sum(sum_votes), 2),
         sum_votes = prettyNum(sum_votes, big.mark = ",")) %>% 
  ungroup %>% 
  arrange(desc(prop_votes))

# inspect
fairfax %>% 
  head -> z
colnames(z) <- c("Electorate","Candidate","Total votes","Proportion votes")
print(xtable::xtable(z), comment=FALSE)
```

We see that the majority of ordinary votes of first preferences in Fairfax actually went to Ted O'Brien, not to Clive Palmer. Palmer's win is due to the preference flow from minor conservative parties, details of this are here: <http://www.tallyroom.com.au/19908>


#Maps and Cartograms

The AEC provides spatial boundaries of Federal Electorates in several formats. The package provides functions for reading these in to R, in order to make maps of the electoral districts and chloropleth maps of results and electoral demographics from the Census data. Figure \ref{results-map} displays the results of the 2016 Federal election on the map. Australia's sparse geography makes it impossible to obtain an accurate impression of the electoral results from this map. 

```{r message=FALSE, error = TRUE, warning=FALSE, echo=FALSE, fig.caption="Electoral results shown on the map. Impossible to get a sense of the results due to the sparseness of the Australian geography."}
data(nat_map_2016)
data(aec2016_fp_electorate)
map.winners <- aec2016_fp_electorate %>% filter(Elected == "Y") %>% 
  select(Electorate, PartyNm) %>% 
  dplyr::rename(Elect_div = Electorate)
  #merge(nat_map_2016, by.x="Electorate", by.y="Elect_div")
# Grouping different Lib/Nats togethers
map.winners$PartyNm <- as.character(map.winners$PartyNm)
coalition <- c("Liberal", "The Nationals")
map.winners.grouped <- mutate(map.winners, 
    GpPartyNm = ifelse(PartyNm %in% coalition,
       "Liberal National Coalition", PartyNm))
map.winners.grouped <- dplyr::left_join(map.winners.grouped,
          distinct(select(nat_map_2016, Elect_div, id)))

# Colour cells to match that parties colours
# Order = Australian Labor Party, Independent, Katters, Lib/Nats Coalition, Xenophon, The Greens
partycolours = c("#FF0033", "#000000", "#CC3300", "#0066CC", "#FF6600", "#009900")
ggplot(data=map.winners.grouped, aes(map_id=id)) + 
  geom_map(aes(fill=GpPartyNm), map=nat_map_2016) +
  scale_fill_manual(name="Political Party", values=partycolours) +
  expand_limits(x=nat_map_2016$long, y=nat_map_2016$lat) +
  theme_map() + coord_equal() + theme(legend.position="bottom")
```

The usual approach to transform small areas on a map to better reflect the distribution of votes is a cartogram (XXX Needs citation). This works very nicely for the USA, where the electorate unit is a state. The geographic area of the states is relatively uniform compared to the Australian electorates' size, and so a cartogram gently expands the small areas, and shrinks the large to enable the gist of the results to be communicated nicely. However, vast open space of western Queensland, New South Wales, South Australia, Western Australia and the Northern Territory relative the the tiny electorates in the city breaks the algorithm, and at very best would so warp the Australian geography to make it unrecognisable. 

The solution we arrived at was to use a Dorling cartogram algorithm which represents geographic areas as circles, and gently shifts the circles to expand important small, close geographic regions. This algorithm is applied separately for each city, and the results are combined. Each electorate becomes a dot on the map, and the dots for city electorates are exploded, to enable each to be seen. Figure \ref{} shows this approach for the 2016 results. 

```{r message=FALSE, error = TRUE, warning=FALSE, echo=FALSE, fig.caption="Electoral results shown on the map. Impossible to get a sense of the results due to the sparseness of the Australian geography."}
data(nat_data_2016_cart)
map.winners.grouped <- dplyr::left_join(map.winners.grouped, select(nat_data_2016_cart, id, x, y))
ggplot(data=map.winners.grouped, aes(map_id=id)) + 
  geom_map(map=nat_map_2016, fill="grey90", colour="white") +
  geom_point(data=map.winners.grouped, aes(x=x, y=y, colour=GpPartyNm),
             alpha=0.6, inherit.aes=FALSE) +
  scale_colour_manual(name="Political Party", values=partycolours) +
  expand_limits(x=nat_map_2016$long, y=nat_map_2016$lat) +
  theme_map() + coord_equal() + theme(legend.position="bottom")
```

#Shiny web app

#Discussion

Discussion about difficulties and futrue development

- what aec and abs could do better
- updating for future elections and censuses
- how this might be used



\bibliography{RJreferences}
